---
title: 'The SAD weekend: a Perilous North American Tradition '
author: "Keri Kemp"
date: "7/20/2020"
output:
  html_document: default
  fig_crop: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview

To determine how diet affects the gut microbiota, pain, and recovery from inflammatory injury, CD1 mice (ICR:Crl) were fed diets varying in fat source: 1) standard chow (REG; NIH-31, Harlan Teklad), 2) standard American Diet (SAD, Envigo TD.140536), 3) Anti-Inflammatory Diet (AID, Envigo TD.150364). A fourth diet group was included to determine the effect of alternating between the diets in which mice were fed the AID diet Monday to Friday and the SAD diet on Saturday and Sunday. This diet shift aimed to model a pattern of eating that included "cheat days" on the weekends with healthy eating during the week. After 14 weeks of diet exposure, when mice were 16 weeks old, a fecal pellet was collected for sequencing of the microbial 16S rRNA gene.

## Raw Data
The original sequencing reads are available for download via NCBI’s Sequence Read Archive under Bioproject PRJNA615647: https://www.ncbi.nlm.nih.gov/sra/PRJNA615647.

``````{r message=FALSE}
#Load libraries
library(tidyverse)
library(phyloseq)
library(vegan)
library(ape)
library(CoDaSeq)
library(zCompositions)
library(Rmisc)
library(agricolae)
library(rcompanion)
library(knitr)
library(kableExtra)
library(cowplot)
library(genefilter)
library(naniar)
library(car)
library(lmtest)
library(ggfortify)
library(indicspecies)
library(pheatmap)
library(btools)
library(RVAideMemoire)
library(data.table)
library(Hmisc)
library(exactRankTests)
library(compositions)
# Set working directory and file paths 
setwd("~/Box/_Sorge_re-analyze")
```

## Import the data into phyloseq

```{r}
# Set file paths 
biom_file = "~/Box/_Sorge_re-analyze/R_input/raw_asv_json.biom"
map_file = "~/Box/_Sorge_re-analyze/R_input/recoverytime_mapping_file.txt"
tree_file = "~/Box/_Sorge_re-analyze/R_input/phylogeny.tre"

# Import biom table, mapping file, and tree file
biom <- import_biom(biom_file, parseFunction = parse_taxonomy_default)
map <- import_qiime_sample_data(map_file)
treefile <- read_tree(tree_file)

# merge in phyloseq object
AID <- merge_phyloseq(biom,map,treefile)

# Inspect the phyloseq object
AID
rank_names(AID)

# Rename columns in the tax_table
tax <- data.frame(tax_table(AID))
tax.clean <- data.frame(row.names = row.names(tax),
Kingdom = str_replace(tax[,1], "D_0__",""),
Phylum = str_replace(tax[,2], "D_1__",""),
Class = str_replace(tax[,3], "D_2__",""),
Order = str_replace(tax[,4], "D_3__",""),
Family = str_replace(tax[,5], "D_4__",""),
Genus = str_replace(tax[,6], "D_5__",""),
Species = str_replace(tax[,7], "D_6__",""),
stringsAsFactors = FALSE)
tax.clean[is.na(tax.clean)] <- ""
tax_table(AID) <- as.matrix(tax.clean)
rank_names(AID)
```


####Prune samples with low sequence count from the dataset
``````{r}
# Prune samples with less than 1,000 reads 
AID = prune_samples(sample_sums(AID)>=1000, AID)
AID
sample_data(AID)

```

####Prune rare taxa and fix non-descriptive taxa
``````{r, results=FALSE, message=FALSE}
# A -- The count value minmum threshold
# k -- The number of samples in which a taxa exceeded A
# A taxa is retained in the dataset if it exceeds the value A in at least k samples.
flist = filterfun(kOverA(2, 5))
AID2 <- filter_taxa(AID, flist, prune=TRUE)
AID2

get_taxa_unique(AID2, "Phylum") #okay
get_taxa_unique(AID2, "Class") #okay
get_taxa_unique(AID2, "Order") #okay
get_taxa_unique(AID2, "Family") 
get_taxa_unique(AID2, "Genus") 
get_taxa_unique(AID2, "Species")
 

# change these non-decriptive labels to NA in order to use the following code to "clean" the taxonomy assignments
# make a list of things to replace with an NA
na_strings <- c("unidentified","uncultured bacterium","gut metagenome","uncultured","mouse gut metagenome", 
                "uncultured rumen bacterium","uncultured bacterium","uncultured organism","human gut metagenome",
                "uncultured Lactobacillus sp.","uncultured Lactobacillus sp.", "uncultured Rheinheimera sp.", 
                "uncultured Lachnospiraceae bacterium","uncultured Clostridiales bacterium","metagenome",
                "uncultured Ruminococcaceae bacterium", "uncultured Bacteroidales bacterium",
                "D\\_3\\_\\_Clostridiales_unclassified","D\\_3\\_\\_Mollicutes_RF39_unclassified",
                "D\\_3\\_\\_Clostridiales_unclassified","D\\_3\\_\\_Mollicutes_RF39_unclassified",
                "D\\_4\\_\\_Lachnospiraceae_unclassified","D\\_4\\_\\_Ruminococcaceae_unclassified",
                "D\\_4\\_\\_Enterobacteriaceae_unclassified","D\\_4\\_\\_Erysipelotrichaceae_unclassified",
                "D\\_5\\_\\_Lachnospiraceae_NK4A136_group_unclassified","D\\_5\\_\\_GCA-900066225_unclassified", 
                "D\\_5\\_\\_Ruminococcaceae_UCG-013_unclassified","D\\_5\\_\\_Caproiciproducens_unclassified",  
                "D\\_5\\_\\_Peptococcus_unclassified","D\\_5\\_\\_Alistipes_unclassified", 
                "D\\_5\\_\\_uncultured_unclassified","D\\_5\\_\\_Coprococcus_3_unclassified",
                "D\\_5\\_\\_Ruminococcaceae_UCG-014_unclassified","D\\_5\\_\\_Pseudomonas_unclassified",
                "D\\_5\\_\\_Acinetobacter_unclassified","D\\_5\\_\\_Streptomyces_unclassified",
                "D\\_5\\_\\_Bacteroides_unclassified","D\\_5\\_\\_Lachnoclostridium_unclassified", 
                "D\\_5\\_\\_Ruminococcus_1_unclassified","D\\_5\\_\\_Odoribacter_unclassified")

# https://cran.r-project.org/web/packages/naniar/vignettes/replace-with-na.html
tax <- data.frame(tax_table(AID2))
tax2<-tax %>%
  rownames_to_column( var = "ASV")  %>%
  naniar::replace_with_na_all(condition = ~.x %in% na_strings) %>%
  column_to_rownames (var = "ASV")

####### Fill holes in the tax table
tax.clean <- tax2
tax.clean[is.na(tax.clean)] <- ""
for (i in 1:nrow(tax.clean)){
  if (tax.clean[i,2] == ""){
    kingdom <- paste("unclassified_", tax.clean[i,1], sep = "")
    tax.clean[i, 2:7] <- kingdom
  } else if (tax.clean[i,3] == ""){
    phylum <- paste("unclassified_", tax.clean[i,2], sep = "")
    tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i,4] == ""){
    class <- paste("unclassified_", tax.clean[i,3], sep = "")
    tax.clean[i, 4:7] <- class
  } else if (tax.clean[i,5] == ""){
    order <- paste("unclassified_", tax.clean[i,4], sep = "")
    tax.clean[i, 5:7] <- order
  } else if (tax.clean[i,6] == ""){
    family <- paste("unclassified_", tax.clean[i,5], sep = "")
    tax.clean[i, 6:7] <- family
  } else if (tax.clean[i,7] == ""){
    tax.clean$Species[i] <- paste("unclassified",tax.clean$Genus[i], sep = "_")
  }
}


tax.clean[is.na(tax.clean)] <- ""
for (i in 1:nrow(tax.clean)){
  if (tax.clean[i,2] == ""){
    kingdom <- paste("Kingdom_", tax.clean[i,1], sep = "")
    tax.clean[i, 2:7] <- kingdom
  } else if (tax.clean[i,3] == ""){
    phylum <- paste("Phylum_", tax.clean[i,2], sep = "")
    tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i,4] == ""){
    class <- paste("Class_", tax.clean[i,3], sep = "")
    tax.clean[i, 4:7] <- class
  } else if (tax.clean[i,5] == ""){
    order <- paste("Order_", tax.clean[i,4], sep = "")
    tax.clean[i, 5:7] <- order
  } else if (tax.clean[i,6] == ""){
    family <- paste("Family_", tax.clean[i,5], sep = "")
    tax.clean[i, 6:7] <- family
  } else if (tax.clean[i,7] == ""){
    tax.clean$Species[i] <- paste("Genus",tax.clean$Genus[i], sep = "_")
  }
}
tax_table(AID2) <- as.matrix(tax.clean)

AID2
# mean, max and min of sample read counts
min(sample_sums(AID2)) #31863
mean(sample_sums(AID2)) #52861.78
max(sample_sums(AID2)) #83610
```

####Fig. 5 Barchart
```{r , message=FALSE,fig.show='hide',results=FALSE}
# Transform to relative abundance
AID_norm <- transform_sample_counts(AID2, function(x) 100 * x/sum(x))

# Prepare to make a barchart at order level
AID_Orders <- AID_norm  %>%
  tax_glom(taxrank = "Order") %>% # agglomerate taxa at order level
  psmelt() %>%                    # Melt phyloseq object to long format for producing graphics with ggplot2
  filter(Abundance > 1.0)  %>%    # Filter out orders below 1% in each sample
  arrange(desc(Phylum),desc(Class),desc(Order))

# Check how many Orders 
sample_data(AID_Orders)$Order #13

# Sum remaining taxa with a relative abundance < 1% and make a new dataframe
Remainders <- (AID_Orders) %>%
  dplyr::group_by(SampleID,Diet,Sex,Cage) %>% 
  dplyr::summarise(Abundance = (100-sum(Abundance))) %>% 
  as.data.frame()
Remainders$Order<-"Orders < 1%"
Remainders$Phylum<-"_Orders < 1%"
Remainders$Class<-"_Orders < 1%"
Remainders$Kingdom<-"Bacteria"

# Compile dataframes
AID_barchart <- full_join(AID_Orders,Remainders)
AID_barchart$Order <- as.factor(AID_barchart$Order)

# Reorder based on Phylogeny
AID_barchart <- AID_barchart[with(AID_barchart, order(Phylum,Class,Order)),]
AID_barchart$Order <- as.factor(AID_barchart$Order)
levels(AID_barchart$Order)
                                                    
# Lock in Order level order
AID_barchart$Order <- factor(AID_barchart$Order, 
      levels = c("Orders < 1%","Bifidobacteriales","Coriobacteriales","Bacteroidales",
                "Deferribacterales","Bacillales","Lactobacillales",
                "Clostridiales","Erysipelotrichales", "Enterobacteriales",
                "Pseudomonadales","Mollicutes RF39","Verrucomicrobiales"))  
                                                            
# Lock in Diet level order
AID_barchart$Diet <- factor(sample_data(AID_barchart)$Diet, levels = c("REG","AID","SAD","AID-SAD"))

write.csv(AID_barchart, file = "R_output/AID_barchart.csv")

# Function for making a custom color palette for plotting
library(RColorBrewer)
# Function for plotting colors side-by-side from here: https://cran.r-project.org/web/packages/colorspace/index.html
pal <- function(col, border = "light gray", ...){
  n <- length(col)
  plot(0, 0, type="n", xlim = c(0, 1), ylim = c(0, 1),
       axes = FALSE, xlab = "", ylab = "", ...)
  rect(0:(n-1)/n, 0, 1:n/n, 1, col = col, border = border)
}

colors2=c("grey", 
          "#0072b2", "#7BAFDE",            #blue
          "#f2a045","#f0e442",             #orange/yellow
          "#117755","#80c175","#CAE0AB",   #green
          "#44AAAA","#9cdada",             #tourqoise
          "#882E72", "#B178A6",            #purple
          "#d55e00")                       #red
pal(colors2)
```

```{r, fig.width = 11, fig.asp = 0.618}
# Plot barchart for Orders for publication
# Export eps as 950 x 500
AID_bc <- ggplot(AID_barchart, aes(x = SampleID, y = Abundance, fill = Order))+
  geom_bar(stat = "identity", colour = "black",size=0.25,width=1)+
  scale_fill_manual(values = colors2,
      labels=c("Orders < 1%","Actinobacteria, Actinobacteria, Bifidobacteriales", 
              "Actinobacteria, Coriobacteriia, Coriobacteriales", 
              "Bacteroidetes, Bacteroidia, Bacteroidales",
              "Deferribacteres, Deferribacteres, Deferribacterales", "Firmicutes, Bacilli, Bacillales", 
              "Firmicutes, Bacilli, Lactobacillales", "Firmicutes, Clostridia, Clostridiales", 
              "Firmicutes, Erysipelotrichi, Erysipelotrichales", "Tenericutes, Mollicutes, RF39",
              "Proteobacteria, Gammaproteobacteria, Enterobacteriales", 
              "Proteobacteria, Gammaproteobacteria, Pseudomonadales", 
              "Verrucomicrobia, Verrucomicrobiae, Verrucomicrobiales"))+
  labs(x = " ",y = "Relative Abundance (%)")+
  labs(fill='Bacterial Order')+
  facet_grid(~Diet, space = "free", scales="free", switch = "x", labeller = labeller(.multi_line = FALSE))+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5, size=14,face="bold"))+
  theme(axis.text.x = element_blank(), axis.ticks.x=element_blank())+
  theme(axis.text.y = element_text(size=11), axis.title.y=element_text(size=12,face="bold"))+
  theme(legend.text=element_text(size=11), legend.title=element_text(size=12,face="bold"))+
  theme(panel.border = element_rect(colour="black", fill=NA, size=0.75, linetype="solid"))+
  theme(strip.text.x = element_text(size=11,face="bold"))+
  theme(plot.margin = unit(c(0.2, 0, 0, 0), "cm"))+
  scale_y_continuous(expand = c(0,0), breaks = seq(0, 100, 10))+
  scale_x_discrete(expand = c(0,0))
AID_bc
```

####Fig 6. Ralative abundance of major Phyla
```{r, fig.show='hide'}
# Agglomerate at Phylum
Phyla <- AID_norm  %>%
  tax_glom(taxrank = "Phylum") %>%                     
  psmelt()   
# Summarize at Phylum with Rmisc function
Phyla_sum <- summarySE(Phyla, measurevar="Abundance", groupvars=c("Phylum","Diet"))
# Export these tables
write.csv(Phyla,"./R_output/Phyla.csv", row.names=FALSE)
write.csv(Phyla_sum,"./R_output/Phyla_sum.csv", row.names=FALSE)

# Filter for stats and graphing
Actino <- Phyla  %>%
  filter(Phylum == "Actinobacteria") 
Bacto <- Phyla  %>%
  filter(Phylum == "Bacteroidetes") 
Firm <- Phyla  %>%
  filter(Phylum == "Firmicutes") 
Proteo <- Phyla  %>%
  filter(Phylum == "Proteobacteria") 

# Set colors (https://www.color-hex.com/color-palette/49436)
cols = c("REG" = "#0072b2", "SAD" = "#d55e00", "AID-SAD" = "#f0e442", "AID" = "#009e73" )

# Re-arrange for plotting
Proteo$Diet <- factor(sample_data(Proteo)$Diet, levels = c("REG","SAD","AID-SAD","AID"))
Proteo_plot <- ggplot(Proteo, aes(x=Diet, y=Abundance,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) + 
  labs(x = " ",y = "Proteobacteria \n Relative Abundance (%)") + theme(legend.position="none") + ylim (0,50) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=14))+
  theme(axis.title.y = element_text(size=16))

Bacto$Diet <- factor(sample_data(Bacto)$Diet, levels = c("REG","SAD","AID-SAD","AID"))
Bacto_plot <- ggplot(Bacto, aes(x=Diet, y=Abundance,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) + 
  labs(x = " ",y = "Bacteroidetes \n Relative Abundance (%)") + theme(legend.position="none") + ylim (0,50) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=14)) +
  theme(axis.title.y = element_text(size=16))

Firm$Diet <- factor(sample_data(Firm)$Diet, levels = c("REG","SAD","AID-SAD","AID"))
Firm_plot <- ggplot(Firm, aes(x=Diet, y=Abundance,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) + 
  labs(x = " ",y = "Firmicutes \n Relative Abundance (%)") + theme(legend.position="none") + ylim (0,80) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=14))+
  theme(axis.title.y = element_text(size=16))

Actino$Diet <- factor(sample_data(Actino)$Diet, levels = c("REG","SAD","AID-SAD","AID"))
Actino_plot <- ggplot(Actino, aes(x=Diet, y=Abundance,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) + 
  labs(x = " ",y = "Actinobacteria \n Relative Abundance (%)") + theme(legend.position="none") + ylim (0,80) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=14))+
  theme(axis.title.y = element_text(size=16))
```

```{r, fig.align='center', fig.width = 9, fig.asp = 0.618}
# Combine Plots
# Export eps at 925w x 750h
theme_set(theme_cowplot())
Phyla_Plot <- plot_grid(Actino_plot, Bacto_plot, Firm_plot, Proteo_plot,  align = 'v', ncol = 2)
Phyla_Plot
```

####Statistics for Figure 6
```{r, fig.align='center', out.width="80%"}
# Histograms
# Create 2x2 plot environment  
par(mfrow = c(2, 2))
# Plot each Metric
plotNormalHistogram((log(Proteo$Abundance)), main= "Proteobacteria")
plotNormalHistogram((log(max(Firm$Abundance+1) - Firm$Abundance)), main= "Firmicutes") 
plotNormalHistogram(Bacto$Abundance, main= "Bacteroidetes")
plotNormalHistogram((log(Actino$Abundance)), main= "Actinobacteria")

# Test for normality 
shapiro.test(log(Proteo$Abundance))
shapiro.test((log(max(Firm$Abundance+1) - Firm$Abundance)))
shapiro.test(Bacto$Abundance) 
shapiro.test(log(Actino$Abundance))
```

#### Actinobacteria
```{r, fig.align='center', out.width="80%"}
####### Actinobacteria #######
lm <-glm(log(Actino$Abundance) ~ Diet*Cage,  data=Actino)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(2), test.statistic=c("F")) #Type II is more powerful when there is no interaction
# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Actinobacteria ANOVA table") %>%
  kable_styling(full_width = FALSE)

# Breusch-Pagan test for heteroskedasticity 
# p>0.05, so we fail to reject the null hypothesis that the variance of the residuals is constant
lmtest::bptest(lm)

# Posthoc multiple comparisons via tukey's HSD test in agricolae
# P-values
agricolae::HSD.test(lm,"Diet", group=TRUE,
                main="Affect of Diet on Actinobacteria", console=TRUE)

# Create 2x2 plot environment
par(mfrow = c(2, 2), mar=c(5,6,4,1)+.1)
# Plot data
boxplot((log(Actino$Abundance)) ~ Diet, data=Actino,
        ylab="Log Transformed \n Relative Abundance (RA) ",
        xlab="Diet", main = "Actinobacteria RA (%)")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles \n for Residuals", main = "Normal Probability Plot")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

#### Bacteroidetes
```{r, fig.align='center', out.width="80%"}
####### Bacteroidetes #######
lm <- glm(Bacto$Abundance ~ Diet*Cage, data=Bacto)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(2), test.statistic=c("F")) #Type II is more powerful when there is no interaction
# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Bacteroidetes ANOVA table") %>%
  kable_styling(full_width = FALSE)

# Breusch-Pagan test for heteroskedasticity 
# p>0.05, so we fail to reject the null hypothesis that the variance of the residuals is constant
lmtest::bptest(lm)

# No post-hoc tests beacuse the global test is not significant)

# Create 2x2 plot environment
par(mfrow = c(2, 2), mar=c(5,6,4,1)+.1)
# Plot data
boxplot(Abundance ~ Diet, data=Bacto,
        ylab="Relative Abundance (RA)",
        xlab="Diet", main = "Bacteroidetes RA (%)")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles \n for Residuals", main = "Normal Probability Plot")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

#### Firmicutes
```{r, fig.align='center', out.width="80%"}
####### Firmicutes #######
lm <- glm((log(max(Firm$Abundance+1) - Firm$Abundance)) ~ Diet*Cage, data=Firm)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(2), test.statistic=c("F")) #Type II is more powerful when there is no interaction
# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Firmicutes ANOVA table") %>%
  kable_styling(full_width = FALSE)

# Breusch-Pagan test for heteroskedasticity 
# p>0.05, so we we fail to reject the null hypothesis that the variance of the residuals is constant
lmtest::bptest(lm)

# No post-hoc tests beacuse the global test is not significant)

# Create 2x2 plot environment
par(mfrow = c(2, 2), mar=c(5,6,4,1)+.1)
# Plot data
boxplot((log(max(Firm$Abundance+1) - Firm$Abundance)) ~ Diet, data=Firm,
        ylab="Relative Abundance (RA)",
        xlab="Diet", main = "Firmicutes RA (%)")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles \n for Residuals", main = "Normal Probability Plot")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

#### Proteobacteria
```{r, fig.align='center', out.width="80%"}
####### Proteobacteria #######
lm <- glm(log(Proteo$Abundance) ~ Diet*Cage, data=Proteo)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(2), test.statistic=c("F")) #Type II is more powerful when there is no interaction
# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Proteobacteria ANOVA table") %>%
  kable_styling(full_width = FALSE)

# Breusch-Pagan test for heteroskedasticity 
# p>0.05, so we fail to reject the null hypothesis that the variance of the residuals is constant
lmtest::bptest(lm)

# Posthoc multiple comparisons via tukey's HSD test in agricolae
# P-values
agricolae::HSD.test(lm,"Diet", group=TRUE,
                main="Affect of Diet on Proteobacteria", console=TRUE)

# Create 2x2 plot environment
par(mfrow = c(2, 2), mar=c(5,6,4,1)+.1)
# Plot data
boxplot((log(Proteo$Abundance)) ~ Diet, data=Proteo,
        ylab="Log Transformed \n Relative Abundance (RA) ",
        xlab="Diet", main = "Proteobacteria RA (%)")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles \n for Residuals", main = "Normal Probability Plot")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

###Fig 7 Beta Diversity, Panel C
####Now export otu and taxa tables from phyloseq for codaseq
```{r}
otu = as(otu_table(AID2), "matrix")
taxon = as(tax_table(AID2), "matrix")
metadata = as(sample_data(AID2), "matrix")
otu = as.data.frame(otu)
alex.in <- otu
taxon = as.data.frame(taxon)
samples = as.data.frame(metadata) # needs to be unique for plotting later
```

####Imput zero values and conduct the Aitchison center-log-ratio (CLR) transformation.
```{r}
# count number of zeros in the table
sum(otu == 0)
# Replace 0 values with an estimate (because normalization is taking log, can't have 0) using the Bayesian-multiplicative replacement function cmultRepl in Zcompositions. Also transposing here, because we need samples as rows.
d.czm <- zCompositions::cmultRepl(t(otu), method="CZM", label=0)
# Perform the centred log-ratio, or CLR (Aitchison) transformation using CoDaSeq.  
# The output will have samples as ROWS
d.clr <- CoDaSeq::codaSeq.clr(d.czm)
```

####Aitchison distance PCA
```{r}
# Samples must be ROWs and features/OTUs as COLUMNS. Use the prcomp included in the R core statistics.
d.pcx <- prcomp(d.clr)
```

```{r, fig.align='center', out.width="60%"}
# Plot PCA with ggplot2 (showing samples only)
df_out <- as.data.frame(d.pcx$x)
theme_set(theme_bw()+theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()))

autoplot(prcomp(d.clr), data = samples, colour = 'Diet')+
  geom_point(aes(color = samples$Diet), size = 3)+ 
  theme(legend.title = element_text(size=14))+theme(legend.text = element_text(size=12))+
  scale_color_manual(values = cols)+
  labs(color="Diet") +
   theme(aspect.ratio = 1, # set plot to square
        axis.title = element_text(size=14), 
        axis.text=element_text(size=12))
```

####Statistics for Fig 7c (betadiversity)
```{r}
# Permanova between groups using Aitchison distance.
dist.clr <- dist(d.clr)
perm<-adonis(dist.clr~Diet*Cage, as(sample_data(AID2),"data.frame"))
print(perm)
```

Define the function: Make knitr table from Permanova results. https://rdrr.io/github/TBrach/MicrobiomeX/src/R/Functions.R
```{r definition, echo = FALSE}
get_permanova_table <- function(adonis) {
  kable_table <- adonis[[1]]
  kable_table[["Pr(>F)"]] <- format(kable_table[["Pr(>F)"]], digits = 3)
  options(knitr.kable.NA = '') 
  knitr::kable(kable_table,
               digits = 3,
               caption = "Permanova results") %>%
  kable_styling(full_width = FALSE)
}
```

```{r}
get_permanova_table
```

Use the function:
```{r}
get_permanova_table(perm)
```

```{r}
# Pairwise adonis
# https:https://www.rdocumentation.org/packages/RVAideMemoire/versions/0.9-69/topics/pairwise.perm.manova
RVAideMemoire::pairwise.perm.manova(dist.clr,samples$Diet,nperm=999,p.method="fdr")
```

```{r}
# Run permdisp to see if statistical differences among communities are driven by differences in variance.
permdisp<-betadisper(dist.clr,samples$Diet)
aov <-anova(permdisp) #this was not significant 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(aov, digits = 3, caption = "Permdisper ANOVA table") %>%
  kable_styling(full_width = FALSE)
```

####Alpha Diversity
```{r, fig.align='center', out.width="80%"}
# Rarefy to account for differences in sequencing depth among samples. Sample without replacement so that the original 
# count value is the maximum possible. Trim OTUs from the dataset that are no longer observed in any sample (have a
# count of zero in every sample).
set.seed(310)
AID_R = rarefy_even_depth(AID2, sample.size = 30000, replace = FALSE, trimOTUs = TRUE)

shannon<-estimate_richness(AID_R, split = TRUE, measures=c("Shannon"))
shannon$SampleID <- rownames(shannon)
# Faith's Phylogenetic Diversity (SR = observed species richness)
FaithsPD<-btools::estimate_pd(AID_R)
colnames(FaithsPD)[1] <-"Faiths_PD"
FaithsPD$SampleID <- rownames(FaithsPD)
# Merge tables
alpha_div <- full_join(shannon,FaithsPD, by = "SampleID")
alpha_div <- alpha_div[,c(2,1,3)]
# Add metadata
alpha_div <- full_join(alpha_div,samples, by = "SampleID")
alpha_div$Cage<-as.numeric(as.character(alpha_div$Cage))
# Write out file
write.csv(alpha_div,"./R_output/alpha_div.csv")

# Summarize for stats and plotting
fpd <- Rmisc::summarySE(alpha_div, measurevar="Faiths_PD", groupvars=c("Diet"))
shan <- Rmisc::summarySE(alpha_div, measurevar="Shannon", groupvars=c("Diet"))
# Write out file
write.csv(fpd,"./R_output/Faiths_PD_Summary.csv")
write.csv(shan,"./R_output/Shannon_Summary.csv")

# Histograms
# Create 2x2 plot environment  
par(mfrow = c(2, 2))
# Plot each Metric
plotNormalHistogram(alpha_div$Faiths_PD, main= "Faiths PD")
plotNormalHistogram(alpha_div$Shannon, main= "Shannon Index")
plotNormalHistogram((log(alpha_div$Faiths_PD)), main= "log(Faiths PD)")

# Test for normality 
shapiro.test(alpha_div$Shannon) 
shapiro.test(log(alpha_div$Faiths_PD)) 
```

####Fig 7 Alpha Diversity, Panels A-C
```{r, fig.align='center', out.width="80%"}
# Re-arrange for plotting
alpha_div$Diet <- factor(sample_data(alpha_div)$Diet, levels = c("REG","SAD","AID-SAD","AID"))

# Set plotting theme 
theme_set(theme_bw())

# Plot Shannon
Shannon_plot <- ggplot(alpha_div, aes(x=Diet, y=Shannon,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) +  
  labs(x = " ",y = "Evenness \n (Shannon Index)") + theme(legend.position="none") + ylim (1,5) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=11)) 

# Plot Faith's PD
PD_plot <- ggplot(alpha_div, aes(x=Diet, y=Faiths_PD,  fill=Diet)) + 
  geom_boxplot(outlier.size = 1) + scale_fill_manual(values = cols) + geom_boxplot(outlier.size = 1) + 
  labs(x = " ",y = "Phylogenetic Diversity \n (Faith's PD)") + theme(legend.position="none") + ylim (5,20) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(text = element_text(size=14),axis.text.x=element_text(size=11)) 

# Combine Plots
theme_set(theme_bw(base_size=14)) # reduce default font size
AlphaDiv_Plot <- plot_grid(Shannon_plot, PD_plot, align = 'v', ncol = 1, axis='l')
AlphaDiv_Plot
```

###Statistics for Fig 7

####Shannon Index
```{r, fig.align='center', out.width="80%"}
lm <-glm(Shannon ~ Diet*Cage, data=alpha_div)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(3), test.statistic=c("F"), contrasts=list(topic=contr.sum, sys=contr.sum))

# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Shannon Index ANOVA table") %>%
  kable_styling(full_width = FALSE) 

# Posthoc multiple comparisons via tukey's HSD test in agricolae
# P-values
agricolae::HSD.test(lm,"Diet", group=FALSE,
                main="Affect of Diet on Shannon Index", console=TRUE)

# Create 2x2 plot environment
par(mfrow = c(2, 2))
# Plot data
boxplot(Shannon ~ Diet, data = alpha_div,
        ylab="Shannon",
        xlab="Diet", main = "Shannon Boxplot")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles for Residuals")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

####Faith's PD
```{r, fig.align='center', out.width="80%"}
lm <-glm((log(Faiths_PD)) ~ Diet*Cage, data=alpha_div)
summary(lm)
car::vif(lm)
sstable <-car::Anova(lm , type = c(3), test.statistic=c("F"), contrasts=list(topic=contr.sum, sys=contr.sum))

# The following code adds a column to the sstable object with partial eta-squared estimates for each effect:
sstable$PES <- c(sstable$'Sum Sq'[-nrow(sstable)], NA)/(sstable$'Sum Sq' + sstable$'Sum Sq'[nrow(sstable)]) 
# Creating an html ANOVA table 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(sstable, digits = 3, caption = "Faith's PD ANOVA table") %>%
  kable_styling(full_width = FALSE)

# Posthoc multiple comparisons via tukey's HSD test in agricolae
# P-values
agricolae::HSD.test(lm,"Diet", group=FALSE,
                main="Affect of Diet on Faiths PD", console=TRUE)

# Create 2x2 plot environment
par(mfrow = c(2, 2))
# Plot data
boxplot((log(Faiths_PD)) ~ Diet, data = alpha_div,
        ylab="PD",
        xlab="Diet", main = "Faiths PD Boxplot")
# Plot residules histogram
x = (residuals(lm))
plotNormalHistogram(x, main = "Residules Histogram")
# Normal probability plot 
qqnorm(residuals(lm), ylab="Sample Quantiles for Residuals")
qqline(residuals(lm), col="red")
# Simple scatterplot between residuals and predicted values. It should look more or less random.
plot(fitted(lm), residuals(lm), main = "Residuals vs. Predicted Values")
```

#####ANCOM-II at the genus level
Download the ANCOM v2.1 script from here: https://github.com/FrederickHuangLin/ANCOM
```{r}
source("~/Box/ANCOM-master/scripts/ancom_v2.1.R")
AID_g = tax_glom(AID2, taxrank = "Genus")
AID_g 
otu = as(otu_table(AID_g), "matrix")
otu_data = as.data.frame(otu)

# Make new datafram for taxa
ID.taxa <- as.data.frame(tax_table(AID_g))
ID.taxa<- rownames_to_column(ID.taxa, var = "taxa_id") 
ID.taxa <- ID.taxa %>% unite("Taxa", 3:7, sep = ", ", remove = FALSE) %>% dplyr::select(taxa_id,Taxa)


metadata = as(sample_data(AID_g), "matrix")
meta_data= as.data.frame(metadata)
meta_data = meta_data %>% dplyr::rename("Sample.ID" = SampleID)
rownames(meta_data) <- c()

# Step 1: Data preprocessing
feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
out_cut = 0.05; zero_cut = 0.90; lib_cut = 1000; neg_lb = FALSE
prepro = feature_table_pre_process(feature_table, meta_data, sample_var, group_var, 
                                   out_cut, zero_cut, lib_cut, neg_lb)
feature_table = prepro$feature_table # Preprocessed feature table
meta_data = prepro$meta_data # Preprocessed metadata
struc_zero = prepro$structure_zeros # Structural zero info

# Step 2: ANCOM
main_var = "Diet"; p_adj_method = "BH"; alpha = 0.05
adj_formula = "Cage"; rand_formula = NULL
t_start = Sys.time()
res = ANCOM(feature_table, meta_data, struc_zero, main_var, p_adj_method, 
            alpha, adj_formula, rand_formula)
t_end = Sys.time()
t_run = t_end - t_start
res_df<-res$out
res2 <- res_df[which(res_df$detected_0.7==TRUE),] 
res2 <- left_join(res2,  ID.taxa, by="taxa_id")
print(res2$Taxa)
```

#####Indispecies at Genus level
```{r}
# Extract abundance matrix from the phyloseq object
OTU1 = as(otu_table(AID_g ), "matrix")
# Coerce to data.frame and transpose (need samples as ROWS and ASVs as COLUMNS)
OTUdf = as.data.frame(t(OTU1))

# Make new datafram for taxa
ID.taxa <- as.data.frame(tax_table(AID_g))
ID.taxa<- rownames_to_column(ID.taxa, var = "ID") 
ID.taxa <- ID.taxa %>% unite("Taxa", 3:7, sep = ", ", remove = FALSE) %>% dplyr::select(ID,Taxa)

# Prepare mapping file
metadata = as(sample_data(AID_g), "matrix")
mapping_file = as.data.frame(metadata)
# First column (SampleID) should be row names
# Remove SampleID column
mapping_file[1] <- NULL
# Make sure the order of the sample names in both match
nrow(mapping_file)==nrow(OTUdf)

# Create factor name to call for Uniqueclass
OTUs.Diet<-as.factor(mapping_file$Diet)
is.factor(OTUs.Diet) #should be TRUE

# Indicator species analysis by unique site*tree identifier
indisp<-multipatt(x=as.data.frame(OTUdf),
                          cluster=OTUs.Diet,
                          func="IndVal.g",
                          duleg=TRUE,
                          #restcomb = c(1,2,3,4,5,9),
                          control=how(nperm=9999))

# Extract table of stats
indisp.sign<-as.data.table(indisp$sign, keep.rownames=TRUE)
# Add adjusted p-value
indisp.sign[ ,p.value.fdr:=p.adjust(p.value, method="fdr")]
# Now can select only the indicators with adjusted significant p-values
tab1<- indisp.sign[p.value.fdr<=0.05, ]
tab1<-tab1%>%dplyr::rename(ID = rn)

# Add taxa information and write out
indicators_genus<- inner_join(tab1, ID.taxa, by ="ID")
write.csv(indicators,"./R_output/indicator_OTUs_summary_wgenus.csv")
print(indicators_genus)

# Get summary information of OTUs that are indicators.
# Note this does not include the FDR adjusted p-values.
print(summary(indisp, alpha=0.05, Bt=0.75 ,indvalcomp=TRUE))

```

####Fig 8 Heat map of bacterial indicators of diet at the genus level
```{r, fig.align='left', fig.width = 12, fig.asp = .62}
AID_g
# log(1+x) transformation
AID_gt <- microbiome::transform(AID_g, 'log10p')
otu = as(otu_table(AID_gt), "matrix")
otu = as.data.frame(otu)
metadata = as(sample_data(AID_gt), "matrix")
samples= as.data.frame(metadata)

# Make new datafram for taxa
ID.taxa <- as.data.frame(tax_table(AID_gt))
ID.taxa<- rownames_to_column(ID.taxa, var = "ID") 
ID.taxa <- ID.taxa %>% unite("Taxa", 3:7, sep = ", ", remove = FALSE) %>% dplyr::select(ID,Taxa)

# Significant ASVs
list <-c("node144" , "node54", "node17" ,  "node55" , "node207", "node162")

# Subset the otu table to include only the otus that were significant indicators
# Add taxonomy
map <- subset(otu, rownames(otu) %in% list)
map <- map %>% rownames_to_column(var = "ID") 
map <- dplyr::inner_join(map, ID.taxa, by ="ID")
map <- map %>% mutate(Taxa = str_replace(Taxa, "Candidatus Arthromitus", "Candidatus"))
# Get ASV names for row annotation
annotation_row <- map %>% dplyr::select(Taxa) 
# Get rid of rownames then reassign to ASV name
rownames(map) <- c()
map <-  map %>% column_to_rownames(var = "Taxa")


# Get column order
sample_order <- samples %>% 
  arrange(factor(Diet, levels = c("SAD", "REG","AID", "AID-SAD"))) %>% 
   dplyr::pull(SampleID)
# Match columns in map to the list in sample_order
map <- map[sample_order]
   
# Set column annotations
annotation_col <- samples %>%
  arrange(factor(Diet, levels = c("SAD", "REG","AID", "AID-SAD"))) %>% 
  column_to_rownames(var = "SampleID") %>%
  dplyr::select(Diet) 

# Set row annotations
annotation_row <- read_csv("annotation_row_genus.csv", col_names = TRUE)
annotation_row <-  annotation_row %>% column_to_rownames(var = "Genus")


# https://sashat.me/2017/01/11/list-of-20-simple-distinct-colors/
# Assign colors
my_colour = list(
  Diet = c("REG" = "#00517f", "SAD" = "#bc5300", "AID-SAD" = "#f0e442", "AID" = "#009e73" ),
  "ANCOM-II" = c("Significant"="#000000", "Not Significant"="#ffffff"),
  "indicspecies" = c("Significant"="#000000", "Not Significant"="#ffffff")
  
)

# Heatmap with pheatmap
# Export eps at 1000x300
hm<-pheatmap(map, clustering_distance_cols = "euclidean", clustering_distance_rows = "euclidean", border_color = "grey60",
             clustering_method = 'complete', fontsize =8, cluster_rows = T, cluster_cols = T,
             annotation_col = annotation_col, annotation_row = annotation_row, annotation_colors = my_colour,
             fontsize_row = 8, fontsize_col = 8, fontsize_number = 8, 
             cellwidth = 10, cellheight = 10, cutree_cols =3, cutree_rows = 2
)
hm
```

```{r}
sessionInfo()
```